---
title: How does human activity affect the movement patterns of wild animals?
subtitle: An analysis of selected datasets from the Movebank animal tracking database
author: Jannis Bolzern and Elke Michlmayr
format:
  html:
    code-fold: true
execute:
  warning: false
  message: false
lang: en
bibliography: bibliography.bib
---

```{r preprocessing}
#| code-summary: preprocessing
``` 

## Abstract

In this project, we aim to investigate how human activity influences the movement patterns of wild animals. Using GPS and Argos tracking data from red foxes, bobcats, and coyotes across rural and remote areas in England, Canada, and the US, we will analyze home range sizes, temporal activity shifts, and habitat selection in relation to human footprint and land use data.


## Introduction

Research questions:

1. **Access to Anthropogenic Food Sources:** Do animals in high human-impact areas exhibit smaller home ranges due to easier access to food resources?
2. **Temporal Shifts in Activity Patterns:** Do animals become more nocturnal in high human-impact areas to avoid direct human encounters?
3. **Habitat Selection in Human-Dominated Landscapes:** How do animals select habitats (e.g., forests, agriculture) under varying levels of human influence?


## Material and Methods

### Datasets

We will use the following wild animal tracking data published on Movebank under a Creative Commons license by three different research groups (see references):
* **Red fox** data from @porteus2024movement for rural areas (GPS-based, Wiltshire, UK)
* Red fox data from @lai2022red for remote areas (Argos-based, from the uninhabitated islands Bylot and Herschel, Canada)
* **Bobcat** and **coyote** data from @prugh2023fear for remote areas with some rural structures (GPS-based, northern Washington, US)

For the human footprint data, we will use the **global 100 meter resolution terrestrial human footprint data (HFP-100)** by @gassert2023operational. The data can be read in Python, R, or any other script that has libraries that can interpret geospatial data (such as folium).

If required, for **land use** in Washington, US, we will rely on the General Land Use Final Dataset published by Washington Spatial Data ([link](https://geo.wa.gov/datasets/a0ddbd4e0e2141b3841a6a42ff5aff46_0/explore?location=48.347066%2C-118.420235%2C9.91)) and for land use in the UK on gov.uk data ([link](https://www.data.gov.uk/dataset/946ce540-de76-441e-bac8-624f30cace8a/land-cover-map-2021-10m-classified-pixels-gb)).

### Data preparation and processing

@laube2011 investigate how temporal scale affects the calculation of movement parameters (speed, sinuosity and turning angle) of animal trajectories. 

Relevant topics:

* Sampling intervals
* Stimmen die Zeiträume überein?
* Daten von verschiedenen Quellen?

HRP-100: use 2020 data and make a subset as quickly as possible

We will use the following libraries:

* readr, tidyr, dplyr library for data processing
* ggplot2 and tmap library for visualization
* sf and terra library for spatial data handling
* move2 and amt library for trajectory handling
* TBD for reading the human footprint data

### Methodology

#### Trajectory Analysis:
Movement paths will be analyzed to identify patterns in speed, direction, and habitat use. Step lengths and turning angles will help infer behavioral states. Step-selection functions (SSFs) will model habitat preferences relative to availability, allowing us to quantify how animals respond to environmental covariates such as human footprint and land use.

#### Home Range Assessment:
Home range sizes will be calculated using kernel density estimation (KDE) and minimum convex polygons (MCPs). This will provide estimates of the area used by each individual. Home ranges will be compared across regions with varying levels of human activity to test whether access to anthropogenic food sources (e.g., garbage, crops) leads to smaller home ranges due to resource concentration.

TBD: There are several ways Wie die home ranges verwendet, es gibt verschiedene Ansätze
Home intervals (could focus on only that)
Zuerst nur für Füchse
Dann auch mit Bobcat und Coyotes
Cross-species analysis

#### Temporal Activity Patterns:
Movement rates will be used to quantify diel activity shifts. We will test whether animals in high human-impact areas exhibit increased nocturnality, potentially as a strategy to avoid direct human encounters. This analysis will reveal how temporal behavior adapts to human presence.

#### Habitat Selection:
SSFs will quantify selection for human-modified habitats (e.g., agricultural areas, urban edges) relative to natural habitats. By comparing selection patterns across species and regions, we will assess how habitat preferences vary with human influence.

## Results

## Discussion

* Wortlimit mit Patrick besprechen
* Runder Bericht in dem die research questions zu Ende beantwortet werden


## Appendix

### Wordcount

<!-- after installing the wordcountadding (see https://github.com/benmarwick/wordcountaddin), remove the line "#| eval: false" -->

```{r}
wordcountaddin::word_count("index.qmd")
```


